Estimating duration using survival analysis

### 2024-03-22 cis-perfect-testing.tex
restructure chapter

% \section{Definitions}
% \todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
% \begin{itemize}
% \item
%   We assume that, for all episodes $i$, $D_i$ is iid and independent
%   of the time of the infection. Define the survival function
%   $\prob(D_i \geq t \mid B_{j(i)} = b, \vec{\theta}) = \prob(D_i \geq t \mid \vec{\theta}) = S_{\vec{\theta}}(t)$,
%   where $\vec{\theta}$ are the parameters controlling the survival
%   distribution (the discussion in this document is valid regardless of
%   the model specified for $S_{\vec{\theta}}$ and hence we consider $\vec{\theta}$
%   as an arbitrary vector of parameters).
% \item
%   The beginning of episode $i$ is known to occur in the interval
%   $[l_{j(i)}^{(b)}, r_{j(i)}^{(b)}]$, and similarly for the end of the infection
%   in $[l_{j(i)}^{(e)}, r_{j(i)}^{(e)}]$.
% \end{itemize}

@@

In this chapter, I develop a method to estimate the duration distribution using CIS data.
The CIS sampling is less frequent, but with has a large sample size (\numprint{4800} detected infection episodes) and a long follow-up.

@@

The model will be used to analyse on the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
This is typically observed, unless the individual is lost to follow-up before their episode is ended, which happens rarely.
I will incorporate episodes excluded in this step in the model for undetected infections.

\subsection{Types of episodes}

Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort and $\nsched$ be the number of distinct test schedules.
A test schedule, $\sched_i$, is the set of days an individual in the cohort is tested at.
Each individual has exactly one test schedule but a test schedule could have multiple individuals associated with it.
A test schedule, $\sched_i$ is the set of all times 
%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Let $i = 1, \dots, \nsched$ index infection schedules and $j = 1, \dots \ntot$ index infection episodes.

All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, \sched_j)$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $\sched_j$ is the set of test times for the individual in which the episode occurs.
The episode's duration is $d_j = e_j - b_j + 1$ (see sec 3.?).
%, the number of days the individual in which the episode occurred, was positive for 
This triplet is assumed iid.

The times $b_j$ and $e_j$ span possible infection and recovery times.
$\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.

Over the short calendar period considered, each individual has zero or one detected episodes.
Let $i \in \set{D}$ indicate $\sched_i$ is associated with a detected episode.
That is, $i \in \set{D}$ if and only if there exists a $j(i)$ such that $b_{j(i)}$ and $e_{j(i)}$ such that $(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode. 
This happens
$(b_{j(i)}, e_{j(i)}, \sched_i)$ is a detected episode when all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_i$ such that $b_{j(i)} \leq t \leq e_{j(i)}$; this condition is equivalent to having at least one positive test for the episode.
    \item $\exists t \in \sched_i$ such that $t < b_{j(i)}$, lower bounding the start of the episode.
    \item $\exists t \in \sched_i$ such that $t > e_{j(i)}$, upper bounding the end of the episode.
\end{enumerate}
For a new context, including recent infections, this condition can be relaxed by considering these episodes as right censored, at little cost except a more complicated likelihood expression.

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_j$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\theta$, discussed in ??.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case, $n_{ia} = 0$ admissible episodes occurred in individuals with test schedule $\sched_i$.
  Otherwise, if $i \in \set{D}$ then $n_{ia} =1$ admissible episode occurred in an individual with test schedule $\sched_{i}$.
  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((b, e) \in \alpha_i \mid \vec{\theta})$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((b, e) \in \Omega_i \mid \vec{\theta})$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected.
  These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((b, e) \in \beta_i \mid \vec{\theta})$.
\end{itemize}

@@

These three classes span all possible episodes and are mutually exclusive.
Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\nsched} n_{ia} + \sum_{i=1}^{\nsched} n_{iu} + 0$.
Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.

Any episode $j$ independently belongs to one of these classes.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, conditional on $\ntot$, the number of episodes in each class, $(n_{ia}, n_{iu}, n_{i\bar{a}})^T$, are distributed multinomially.
with probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), which are conditional on these episode occurring with test schedule $\sched_i$.
By multiplying these probabilities by $\eta_i/\Ncis$, where $\eta_i$ is the number of individuals with test schedule $\sched_i$, the probability that an episode belongs to each class, without conditioning on a specific individual can be derived.

{\color{red} Would this be the place to introduce a new subsection? on how to calculate the likelihood? Or on line 242?}
The undetected classes for each individual cannot be distinguished by any observable data.
Each class's count needs to be marginalised over to calculate the likelihood.
Because the counts have a joint multinomial distribution, this is equivalent to only considering the sum of the counts.
The overall probability of an episode being unobserved is $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\nsched} \eta_i p_{iu}$.
Furthermore, the inadmissible classes are guaranteed to have zero episodes.
Counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
Therefore, I combine the undetected classes into a single class and ignore the inadmissible classes.

{\color{red} subsection here instead of line 233?}
Let $\na = (n_{1a}, \dots, n_{\nsched a})^T$ be the vector of counts of episodes in the admissible classes.
The counts $\na$ and $\nnodet$ are jointly distributed multinomially.
The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\nsched} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
Hence:
\begin{align}
p(\na \mid \ntot, \theta)
&= p(\nnodet, \na\mid \ntot, \theta) \\
&= \frac{\ntot!}{\nnodet! \prod_{i=1}^{\nsched} n_{ia}!}  \pnodet^{\nnodet}\prod_{i=1}^{\nsched} \left(\frac{1}{\nsched}p_{ia}\right)^{n_{ia}} \\
&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
\end{align}
the final line follows because $\nnodet = \ntot - \ndet$, $n_{ia} = 1$ for $i \in \set{D}$, and $n_{ia} = 0$ otherwise.

In forming the probability $p_{ia}$, I use the test outcomes.
This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own cell is possible.
The new cells would align with the inadmissible region.
Since each of these cells have an observed count of 0, they can be combined to give the same likelihood.
That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
Therefore, I present this version as it is clearer.

The posterior density we are interested in is $p(\theta \mid \na)$.
It can be written as follows.
\begin{align}
p(\theta \mid \na)
&\propto p(\theta) p(\na \mid \theta) \\
&= p(\theta) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) p(\na \mid \theta, \ntot) \\
&\propto p(\theta) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
&= p(\theta) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:full-posterior}.
\end{align}
The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.

To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \theta) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
I assume that $\theta$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
Therefore, $p(\ntot \mid \theta) = p(\ntot)$.
Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

Substituting this into \cref{perf-test:eq:full-posterior} gives:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior-simplified}.
\end{align}

@@

Next, I derive $p_{ia}$.
By definition (see~\cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((b, e) \in \alpha_i \mid \vec{\theta}) \\
\alpha_i =& \{ (b, e) : l_{j(i)}^{(b)} \leq b \leq r_{j(i)}^{(b)} \wedge l_{j(i)}^{(e)} \leq e \leq r_{j(i)}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$, this gives:}
p_{ia}
=& \prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)}, l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \right) \\
=& \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
   &\times\prob \left( l_{j(i)}^{(b)} \leq B_{j(i)} \leq r_{j(i)}^{(b)} \right) \\
=& \sum_{b = l_{j(i)}^{(l)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} \leq E_{j(i)} \leq r_{j(i)}^{(e)} \mid B_{j(i)} = b \right) \prob \left(B_{j(i)} = b \right) \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \prob \left( l_{j(i)}^{(e)} - b + 1 \leq D_{j(i)} \leq r_{j(i)}^{(e)} - b + 1 \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $D_{j(i)}$} \\
=& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right) \prob \left(B_{j(i)} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j(i)}^{(b)}}^{r_{j(i)}^{(b)}} \left( S_{\vec{\theta}}(l_{j(i)}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j(i)}^{(e)} - b + 1) \right)
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data with truncation.

>>>>

% \section{Definitions}
% \todo[inline]{These will be defined in the introductory chapters, but are here so that this draft is understandable}
% \begin{itemize}
% \item
%   We assume that, for all episodes $i$, $D_i$ is iid and independent
%   of the time of the infection. Define the survival function
%   $\prob(D_i \geq t \mid B_{j} = b, \vec{\theta}) = \prob(D_i \geq t \mid \vec{\theta}) = S_{\vec{\theta}}(t)$,
%   where $\vec{\theta}$ are the parameters controlling the survival
%   distribution (the discussion in this document is valid regardless of
%   the model specified for $S_{\vec{\theta}}$ and hence we consider $\vec{\theta}$
%   as an arbitrary vector of parameters).
% \item
%   The beginning of episode $i$ is known to occur in the interval
%   $[l_{j}^{(b)}, r_{j}^{(b)}]$, and similarly for the end of the infection
%   in $[l_{j}^{(e)}, r_{j}^{(e)}]$.
% \end{itemize}

@@

The CIS sampling is less frequent, but with has a large sample size (\numprint{4800} detected infection episodes) and a long follow-up.
Therefore, it is suitable for estimating the tail of the duration distribution.
In this chapter, I develop a method to estimate the duration distribution using CIS data, assuming no false negatives or positives.

@@

The model will be used to analyse the 4,800 episodes first detected between 10th Oct 2020 and 6th Dec 2020 inclusive, with negatives bounding the start and end time of the episode.
This period is denoted as $t = 1, \dots, T$.
A negative test before the start of the episode provides a lower bound on the episode's start time; otherwise, there is little information on its length.
Similar information comes from the negative test after the end of the episode.
This is typically observed, unless the individual is lost to follow-up before their episode is ended, which happens rarely.

\subsection{Types of episodes} \label{perf-test:sec:episode-types}

Let the total number of infection episodes in the cohort be $\ntot = \ndet + \nnodet$ where $\ndet$ is the number of detected infections and $\nnodet$ is the number of undetected infections.
Let $\Ncis$ be the number of individuals in the cohort.
%Without loss of generality, let the individuals with a detected episode be indexed by $i = 1, \dots, \ndet$ and the individuals without a detected episode be indexed by $i = \ndet + 1, \dots, \Ncis$.
Let $i = 1, \dots, \Ncis$ index individuals and $j = 1, \dots \ntot$ index infection episodes.
The test schedule $\sched_i$ is the set of times individual $i$ is tested at, from their last test prior to time 1 (if it exists) up until time $T$.
This is all the times the infection episode could be detected at and be included, plus one prior test.
For example, for the individual in \cref{perf-test:sec:undetected}, $\sched_i = \{ 0, 7, 14, 21, 28, 56 \}$.
Each individual has exactly one test schedule, but a test schedule could have multiple individuals associated with it.

All the relevant information about the infection episode $j$ can be fully characterized by the (unobserved) triplet $(b_j, e_j, i(j))$ where $b_j$ and $e_j$ are the times the episode begin and end respectively and $i(j)$ denotes the individual the episode occurred in.
The episode's duration is $d_j = e_j - b_j + 1$ (see \cref{E-inc-prev:sec:definitions}).
%, the number of days the individual in which the episode occurred, was positive for 
% This triplet is assumed iid.

% The times $b_j$ and $e_j$ span possible infection and recovery times.
% $\sched_j$ is one of the observed test schedules, with any other test schedule assumed to occur with probability zero.

Over the short calendar period considered, each individual has zero or one detected episodes.
Let $i \in \set{D}$ indicate individual $i$ has a detected episode.
I define the precise conditions for this to occur in \cref{perf-test:sec:prob-undetected}.
% That is, $i \in \set{D}$ if and only if there exists a $j$ such that $b_{j}$ and $e_{j}$ such that $(b_{j}, e_{j}, \sched_i)$ is a detected episode.

To relate the possible value for $b_j$ and $e_j$, given a test schedule $\sched_{i(j)}$, to the data, I define three subsets partitioning the space of possible infection and recovery times (\ie: are strict subsets of $T \times T$).
Figure \ref{perf-test:fig:partitionSpace} shows this graphically.
Each episode, whether detected or not, must fall into one of these three classes with a probability that is a function of the parameters of the survival function, $\vec{\vec{\theta}}$, discussed in \cref{perf-test:sec:parameters-priors}.

\begin{itemize}
\item
  Admissible episodes, $\alpha_i$, which have an infection and recovery time consistent with the data observed.
  This space is empty for $i \notin \set{D}$; in this case individual $i$ had $n_{ia} = 0$ admissible episodes.
  Otherwise, if $i \in \set{D}$, then individual $i$ had $n_{ia} =1$ admissible episode.
  Conditional on the episode occurring in individual $i$, the episode is admissible with probability $p_{ia} = \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i)$.
  \label{perf-test:def:admissible}
\item
  Undetected episodes, $\Omega_i$, which have an infection and recovery time such that they would not have tested positive.
  An unknown number, $n_{iu}$ of these occur in individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in individuals with test schedule $\sched_i$, the episode is undetected with probability $p_{iu} = \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
\item
  Inadmissible episodes, $\beta_i$, which are not consistent with the data observed but would have been
  detected.
  These are all remaining episodes not in the previous sets, with $n_{i\bar{a}} = 0$ of these occurred in any individuals with test schedule $\sched_i$.
  Conditional on the episode occurring in these individuals, the episode inadmissible with probability $p_{i\bar{a}} = \prob((B_j, E_j) \in \beta_i \mid \vec{\theta}, i(j) = i)$.
\end{itemize}

@@

These three classes span all possible episodes and are mutually exclusive.
Hence, $\ntot = \ndet + \nnodet + n_{i\bar{a}} = \sum_{i=1}^{\Ncis} n_{ia} + \sum_{i=1}^{\Ncis} n_{iu} + 0$.
Furthermore, for all $i$, $p_{ia} + p_{iu} + p_{i\bar{a}} = 1$.

% Any episode $j$ independently belongs to one of these classes.
The probabilities ($p_{iu}$, $p_{ia}$, and $p_{i\bar{a}}$), are conditional on the episode occurring in individual $i$.
I assume the individual in which each episode occurs is iid uniform across the cohort.
Therefore, the unconditional probabilities of an infection belonging to a class need to be multiplied by $1/\Ncis$.
For example, the probability of an episode being in class $\alpha_i$ is $p_{ia} / \Ncis$.

Each episode occurs in one of the classes independently and with the same probability.
Therefore, conditional on $\ntot$, the number of episodes in each class is multinomial.
Specifically:
\begin{align}
  \begin{bmatrix}
    n_{1a} \\ n_{1u} \\ n_{1\bar{a}} \\ \vdots \\ n_{\Ncis,a} \\ n_{\Ncis,u} \\ n_{\Ncis,\bar{a}}
  \end{bmatrix}
  \mid \ntot
  =
  \MN\left(
    \ntot,
    \frac{1}{\Ncis}
    \begin{bmatrix}
      p_{1a} \\ p_{1u} \\ p_{1\bar{a}} \\ \vdots \\ p_{\Ncis,a} \\ p_{\Ncis,u} \\ p_{\Ncis,\bar{a}}
    \end{bmatrix}
  \right)
\end{align}

In forming the region $\alpha_i$, I use the test outcomes.
This is not strictly correct because the cell probabilities in the likelihood should not depend on the data.
However, an alternative where each combination of infection and recovery times is considered its own cell is possible.
The new cells would align with the inadmissible region.
Since each of these cells have an observed count of 0, they can be combined to give the same likelihood.
That is, the approach is mathematically equivalent but greatly complicates the formulae and notation.
Therefore, I present this version as it is clearer.

The undetected classes cannot be distinguished.
Therefore, these classes can be combined into a single class.
The combined class has a count $\nnodet = \sum_{i=1}^{\Ncis} n_{iu}$ and probability $\pnodet = \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu}$.

\subsection{Posterior density}

Let $\na = (n_{1a}, \dots, n_{\Ncis a})^T$ be the vector of counts of episodes in the admissible classes.
This represents all our observations.
Therefore, the posterior density we are interested in is $p(\vec{\theta} \mid \na)$.
It can be written as follows.
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) p(\na \mid \vec{\theta}) \\
&= p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) p(\na \mid \ntot, \vec{\theta})
\label{perf-test:eq:posterior1}
\end{align}
where the last line augments the likelihood with the total number of episodes, $\ntot$.

From \cref{perf-test:sec:episode-types}, counts $\na$ and $\nnodet$ conditional on $\ntot$ are jointly distributed multinomially.
All $n_{i\bar{a}} = 0$, therefore these classes can be ignored; counts of zero do not contribute to the multinomial probability mass function (they multiply by 1).
The support of the distribution requires $\nnodet = \ntot - \sum_{i=1}^{\Ncis} n_{ia} = \ntot - \ndet$ and $\ntot \geq \ndet$.
Hence:
\begin{align}
p(\na \mid \ntot, \vec{\theta})
&= p(\nnodet  = \ntot - \na, \na\mid \ntot, \vec{\theta}) \\
&= \frac{\ntot!}{(\ntot - \ndet)! \prod_{i=1}^{\Ncis} n_{ia}!}  \pnodet^{\ntot - \ndet}\prod_{i=1}^{\Ncis} \left(\frac{1}{\Ncis}p_{ia}\right)^{n_{ia}} \\
&\propto \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \label{perf-test:eq:augmented-likelihood}
\end{align}
the final line follows because $n_{ia} = 1$ for $i \in \set{D}$ and $n_{ia} = 0$ otherwise.

Substituting \cref{perf-test:eq:augmented-likelihood} into \cref{perf-test:eq:posterior1}:
\begin{align}
&\propto p(\vec{\theta}) \sum_{\ntot=\ndet}^\infty \left( p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \prod_{i \in \set{D}} p_{ia} \right) \\
&= p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) \left( \sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet} \right) \label{perf-test:eq:posterior2}.
\end{align}
The rest of this section derives expressions for each $p_{ia}$, $p_{u}$ and the sum.

\subsection{Expression for the sum}

To start, I derive an analytical solution to $\sum_{\ntot=\ndet}^\infty p(\ntot \mid \vec{\theta}) \frac{\ntot!}{(\ntot - \ndet)!} \pnodet^{\ntot - \ndet}$ under the prior $\ntot \dist \NBc(\mu, r)$.
I assume that $\vec{\theta}$, which are only the parameters of the survival distribution, has a prior independent of $\ntot$, the number of infections that occurred in the cohort.
Therefore, $p(\ntot \mid \vec{\theta}) = p(\ntot)$.
Putting a negative binomial prior on $\ntot$ is equivalent to the following gamma-Poisson composite; its use simplifies the derivation.
\begin{align}
\ntot \mid \lambda &\dist \Poi(\lambda) \\
\lambda &\dist \GamDist(a, b)
\end{align}
where $b = r / \mu$ and $a = r$.
Hence:
\begin{align}
&\sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot) \\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} p(\ntot \mid \lambda) p(\lambda) d\lambda &\text{$\lambda$ explicit}\\
&= \int \sum_{\ntot=\ndet}^\infty \frac{\ntot!}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \frac{\lambda^{\ntot} e^{-\lambda}}{\ntot!} p(\lambda) d\lambda &\ntot \dist \Poi\\
%&= \int \sum_{\ntot=\ndet}^\infty \frac{1}{(\ntot-\ndet)!} \pnodet^{\ntot-\ndet} \lambda^{\ntot-\ndet} \lambda^{\ndet} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) \sum_{\nnodet=0}^\infty \frac{(\pnodet \lambda)^{\nnodet}}{\nnodet!} d\lambda &\nnodet = \ntot-\ndet\\
&= \int \lambda^{\ndet} e^{-\lambda} p(\lambda) e^{\lambda \pnodet} d\lambda &\text{Maclaurin series of $e$} \\
&= \int \lambda^{\ndet} e^{-\lambda(1 - \pnodet)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda &\text{Gamma pdf}\\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+\ndet-1} e^{-(b+1-\pnodet)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+\ndet)}{(b+1-\pnodet)^{a+\ndet}} &\text{Gamma pdf}\\
&\propto (b+1-\pnodet)^{-(a+\ndet)} \\
&= (r/\mu + 1 - \pnodet)^{-(r+\ndet)} &\text{sub in $\mu$ and $r$}\\
&\propto(r + \mu (1- \pnodet))^{-(r+\ndet)}.
\end{align}

Substituting this into \cref{perf-test:eq:posterior2} gives:
\begin{align}
p(\vec{\theta} \mid \na)
&\propto p(\vec{\theta}) \left( \prod_{i \in \set{D}} p_{ia} \right) (r + \mu (1- \pnodet))^{-(r+\ndet)} \label{perf-test:eq:full-posterior}.
\end{align}

@@

\subsection{Admissible probabilities}

Next, I derive $p_{ia}$.
From \cref{perf-test:eq:full-posterior}, this is only needed for $i \in \set{D}$, that is for individuals with detected episodes.
For such an individual, let $j$ denote their detected episode.

By definition (see \cpageref{perf-test:def:admissible}), we have, for any $i \in \set{D}$:
\begin{align}
p_{ia} =& \prob((B_j, E_j) \in \alpha_i \mid \vec{\theta}, i(j) = i) \\
\alpha_i =& \{ (b, e) : l_{j}^{(b)} \leq b \leq r_{j}^{(b)} \wedge l_{j}^{(e)} \leq e \leq r_{j}^{(e)}\}.
\intertext{Omitting the conditioning on $\vec{\theta}$ and $i(j) = i$, this gives:}
p_{ia}
=& \prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)}, l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \right) \\
=& \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
   &\times\prob \left( l_{j}^{(b)} \leq B_{j} \leq r_{j}^{(b)} \right) \\
=& \sum_{b = l_{j}^{(l)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} \leq E_{j} \leq r_{j}^{(e)} \mid B_{j} = b \right) \prob \left(B_{j} = b \right) \\
=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \prob \left( l_{j}^{(e)} - b + 1 \leq D_{j} \leq r_{j}^{(e)} - b + 1 \right) \prob \left(B_{j} = b \right) &\text{by def of $D_{j}$} \\
=& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 1) \right) \prob \left(B_{j} = b \right) &\text{by def of $S_{\vec{\theta}}$} \\
\propto& \sum_{b = l_{j}^{(b)}}^{r_{j}^{(b)}} \left( S_{\vec{\theta}}(l_{j}^{(e)} - b + 1) - S_{\vec{\theta}}(r_{j}^{(e)} - b + 1) \right)
\end{align}
under the assumption of uniform probability of infection time.
This is the standard form of the likelihood for doubly interval censored data without truncation.

\subsection{Undetected probability} \label{perf-test:sec:prob-undetected}

The final component of \cref{perf-test:eq:full-posterior} required is $1 - p_u$.
From \cref{perf-test:sec:episode-types}:
\begin{align}
  1 - p_u
  &= 1 - \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} p_{iu} \\
  &= \frac{1}{\Ncis} \sum_{i=1}^{\Ncis} (1 - p_{iu}).
  \label{perf-test:eq:pu}
\end{align}

Therefore, the crucial component is $1 - p_{iu} = 1 - \prob((B_j, E_j) \in \Omega_i \mid \vec{\theta}, i(j) = i)$.
This is one minus the probability that an episode in individual $i$ was undetected, \ie the probability that episode such an episode is detected.

An episode $j$ in individual $i(j)$ is detected if and only if all the following conditions are met.
\begin{enumerate}
    \item $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$; this condition is equivalent to having at least one positive test for the episode.
    \item $\exists t \in \sched_{i(j)}$ such that $t < b_j$.
      For individuals enrolled during the period considered ($\min \sched_{i(j)} > 0$), this ensures that the beginning of the episode is lower bounded.
      For individuals enrolled prior to the period considered ($\min \sched_{i(j)} \leq 0$), this means that the episode was not detected prior to time 1.
    \item $\exists t \in \sched_{i(j)}$ such that $t > e_j$, upper bounding the end of the episode.
      For episodes detected in the period I consider, this happens a negligible number of times, only in individuals lost to follow-up; therefore, I neglect this possibility.
      For a new context, including recent infections, this condition could be relaxed by considering episodes that do not meet this criteria as right censored.
\end{enumerate}

% An episode is undetected if and only if no tests are performed during the episode or if there was no negative test prior to the episode.
% Equivalently, that the first test at or after $b$ is after $e$, or that there is no negative test prior to $b$.
To formalise this, first define $\tau_{\sched_i}(t)$ as the time until the next test at or after time $t$ in the schedule $\sched_i$:
\begin{align}
\tau_{\sched_i}(t) &= \min \{ t' \in \sched_i : t' \geq t \} - t.
\end{align}
The condition that $\exists t \in \sched_{i(j)}$ such that $b_j \leq t \leq e_j$ can now be expressed as $e_j \geq b_j + \tau_{\sched_{i(j)}}(b_j)$.
Equivalently, $d_j \geq \tau_{\sched_{i(j)}}(b_j) + 1$.
